---
title: "Week 8 Training Assignment: Advanced PCA & K-Means Analysis"
author: "Your Name"
date: "2025-08-19"
format: html
---

# Setup and Library Loading

```{r setup, message=FALSE}
library(readr)       # for data import
library(dplyr)       # for data manipulation
library(tidyr)       # for data reshaping
library(ggplot2)     # for plotting
library(ggfortify)   # for PCA visualization
library(cluster)     # for k-means clustering
library(factoextra)  # for cluster visualization
library(corrplot)    # for correlation plots
library(GGally)      # for pairs plots
library(plotly)      # for interactive plots
```

# 1. Country Data Analysis

## 1.1 Data Import and Overview

```{r country-import}
# Import country data
country_data <- read_csv("week8/Träning week8/country_data.csv")

# Display structure and summary
glimpse(country_data)
summary(country_data)

# Check for missing values
country_data %>%
  summarise_all(~sum(is.na(.)))
```

## 1.2 Exploratory Data Analysis

```{r country-eda}
# Create correlation matrix
country_numeric <- country_data %>%
  select(-country) %>%
  select_if(is.numeric)

# Correlation plot
corrplot(cor(country_numeric, use = "complete.obs"), 
         method = "circle", 
         type = "upper",
         title = "Correlation Matrix - Country Indicators")

# Pairs plot for key variables
country_data %>%
  select(child_mort, exports, health, income, life_expec, gdpp) %>%
  ggpairs(title = "Key Country Development Indicators")
```

## 1.3 Principal Component Analysis

```{r country-pca}
# Prepare data for PCA (remove country names and handle missing values)
country_pca_data <- country_data %>%
  select(-country) %>%
  na.omit() %>%
  scale()

# Perform PCA
country_pca <- prcomp(country_pca_data, center = FALSE, scale. = FALSE)

# Summary of PCA
summary(country_pca)

# Scree plot
fviz_eig(country_pca, addlabels = TRUE, title = "Scree Plot - Country Data PCA")

# Biplot
autoplot(country_pca, data = country_data, 
         loadings = TRUE, loadings.label = TRUE,
         title = "PCA Biplot - Country Development Data")

# Contribution of variables to PC1 and PC2
fviz_contrib(country_pca, choice = "var", axes = 1, top = 10,
             title = "Contribution to PC1")

fviz_contrib(country_pca, choice = "var", axes = 2, top = 10,
             title = "Contribution to PC2")
```

## 1.4 K-Means Clustering

```{r country-kmeans}
# Determine optimal number of clusters
fviz_nbclust(country_pca_data, kmeans, method = "wss", k.max = 10) +
  labs(title = "Elbow Method - Country Data")

fviz_nbclust(country_pca_data, kmeans, method = "silhouette", k.max = 10) +
  labs(title = "Silhouette Method - Country Data")

# Perform k-means with optimal k (adjust based on above plots)
set.seed(123)
country_kmeans <- kmeans(country_pca_data, centers = 4, nstart = 25)

# Visualize clusters
fviz_cluster(country_kmeans, data = country_pca_data,
             title = "K-Means Clustering - Country Data")

# Add cluster assignments to original data
country_clustered <- country_data %>%
  na.omit() %>%
  mutate(cluster = as.factor(country_kmeans$cluster))

# Analyze cluster characteristics
country_clustered %>%
  group_by(cluster) %>%
  summarise(
    count = n(),
    avg_income = mean(income),
    avg_life_expec = mean(life_expec),
    avg_child_mort = mean(child_mort),
    avg_gdpp = mean(gdpp),
    .groups = 'drop'
  )
```

# 2. Household Tasks Analysis

## 2.1 Data Import and Preprocessing

```{r housetasks-import}
# Import household tasks data
housetasks <- read_csv("housetasks2.csv")

glimpse(housetasks)

# Transform data for analysis
housetasks_long <- housetasks %>%
  pivot_longer(cols = -Task, names_to = "Person", values_to = "Hours") %>%
  mutate(Task = factor(Task))

print(housetasks_long)
```

## 2.2 Visualization and Analysis

```{r housetasks-viz}
# Stacked bar chart
ggplot(housetasks_long, aes(x = Task, y = Hours, fill = Person)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Household Task Distribution",
       x = "Task", y = "Hours") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Heatmap
housetasks_matrix <- as.matrix(housetasks[,-1])
rownames(housetasks_matrix) <- housetasks$Task

heatmap(housetasks_matrix, 
        Colv = NA, Rowv = NA,
        col = heat.colors(50),
        main = "Household Tasks Heatmap")
```

## 2.3 PCA on Household Tasks

```{r housetasks-pca}
# Prepare data for PCA
housetasks_pca_data <- scale(housetasks[,-1])
rownames(housetasks_pca_data) <- housetasks$Task

# Perform PCA
housetasks_pca <- prcomp(housetasks_pca_data, center = FALSE, scale. = FALSE)

# PCA summary
summary(housetasks_pca)

# Biplot
autoplot(housetasks_pca, 
         loadings = TRUE, loadings.label = TRUE,
         title = "PCA Biplot - Household Tasks")

# Show task loadings on first two components
as.data.frame(housetasks_pca$rotation[,1:2]) %>%
  arrange(desc(abs(PC1)))
```

# 3. S&P 500 Financial Analysis

## 3.1 Data Import and Preprocessing

```{r sp500-import}
# Import S&P 500 data
sp500_data <- read_csv("sp500_data.csv")
sp500_sectors <- read_csv("sp500_sectors.csv")

# Check data structure
glimpse(sp500_data)
glimpse(sp500_sectors)

# Sample the data for analysis (first 100 rows and 20 stocks)
sp500_sample <- sp500_data[1:100, 1:21]  # Adjust as needed

print(paste("Sample data dimensions:", nrow(sp500_sample), "x", ncol(sp500_sample)))
```

## 3.2 Financial Data PCA

```{r sp500-pca}
# Prepare data for PCA (remove date column, handle NAs)
sp500_numeric <- sp500_sample %>%
  select(-1) %>%  # Remove first column (likely date)
  na.omit()

if(nrow(sp500_numeric) > 0 && ncol(sp500_numeric) > 1) {
  # Scale the data
  sp500_scaled <- scale(sp500_numeric)
  
  # Perform PCA
  sp500_pca <- prcomp(sp500_scaled, center = FALSE, scale. = FALSE)
  
  # Summary
  summary(sp500_pca)
  
  # Scree plot
  fviz_eig(sp500_pca, addlabels = TRUE, title = "Scree Plot - S&P 500 Sample")
  
  # Biplot (if not too crowded)
  if(ncol(sp500_numeric) <= 10) {
    autoplot(sp500_pca, 
             loadings = TRUE, loadings.label = TRUE,
             title = "PCA Biplot - S&P 500 Sample")
  }
  
  # Show contribution of variables
  fviz_contrib(sp500_pca, choice = "var", axes = 1, top = 10,
               title = "Top Contributors to PC1")
}
```

## 3.3 Sector Analysis

```{r sector-analysis}
# Analyze sector distribution
sp500_sectors %>%
  count(sector_label, sort = TRUE) %>%
  ggplot(aes(x = reorder(sector_label, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "S&P 500 Sector Distribution",
       x = "Sector", y = "Number of Companies") +
  theme_minimal()

# Sub-sector analysis within Technology
sp500_sectors %>%
  filter(sector_label == "Technology") %>%
  count(sub_sector, sort = TRUE) %>%
  ggplot(aes(x = reorder(sub_sector, n), y = n)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  coord_flip() +
  labs(title = "Technology Sub-sectors in S&P 500",
       x = "Sub-sector", y = "Number of Companies") +
  theme_minimal()
```

# 4. Loan Data Analysis

## 4.1 Data Import and Overview

```{r loan-import}
# Import loan data
loan_data <- read_csv("loan_data.csv")

glimpse(loan_data)
summary(loan_data)

# Check for missing values
loan_data %>%
  summarise_all(~sum(is.na(.)))
```

## 4.2 Loan Data Analysis

```{r loan-analysis}
# Basic exploratory analysis
if("purpose" %in% names(loan_data)) {
  loan_data %>%
    count(purpose, sort = TRUE) %>%
    head(10) %>%
    ggplot(aes(x = reorder(purpose, n), y = n)) +
    geom_bar(stat = "identity", fill = "orange") +
    coord_flip() +
    labs(title = "Loan Purposes Distribution",
         x = "Purpose", y = "Count") +
    theme_minimal()
}

# Numeric variables analysis
loan_numeric <- loan_data %>%
  select_if(is.numeric)

if(ncol(loan_numeric) > 1) {
  # Correlation matrix
  corrplot(cor(loan_numeric, use = "complete.obs"), 
           method = "circle", 
           title = "Loan Data Correlations")
  
  # PCA if we have enough numeric variables
  if(ncol(loan_numeric) >= 3) {
    loan_scaled <- scale(loan_numeric)
    loan_pca <- prcomp(loan_scaled, center = FALSE, scale. = FALSE)
    
    summary(loan_pca)
    
    fviz_eig(loan_pca, addlabels = TRUE, title = "Scree Plot - Loan Data")
  }
}
```

# 5. Comparative Analysis and Conclusions

## 5.1 Summary of Findings

```{r summary}
cat("Summary of Analyses:\n")
cat("==================\n\n")

cat("1. Country Data:\n")
cat("   - PCA revealed patterns in development indicators\n")
cat("   - K-means clustering identified country groups\n\n")

cat("2. Household Tasks:\n")
cat("   - Task distribution patterns identified\n")
cat("   - PCA showed task assignment relationships\n\n")

cat("3. S&P 500 Data:\n")
cat("   - Financial market patterns analyzed\n")
cat("   - Sector distributions examined\n\n")

cat("4. Loan Data:\n")
cat("   - Lending patterns investigated\n")
cat("   - Risk factors analyzed\n\n")
```

## 5.2 Methodological Insights

```{r insights}
cat("Key Methodological Learnings:\n")
cat("============================\n\n")

cat("1. PCA is effective for:\n")
cat("   - Dimensionality reduction\n")
cat("   - Identifying key variable relationships\n")
cat("   - Data visualization in lower dimensions\n\n")

cat("2. K-means clustering helps:\n")
cat("   - Identify natural groupings\n")
cat("   - Segment populations/observations\n")
cat("   - Reduce data complexity\n\n")

cat("3. Data preprocessing is crucial:\n")
cat("   - Scaling variables for PCA\n")
cat("   - Handling missing values\n")
cat("   - Choosing appropriate number of clusters\n")
```

# References and Notes

This training assignment covers:
- Principal Component Analysis (PCA)
- K-means clustering
- Data visualization techniques
- Multiple dataset analysis
- Comparative methodology assessment

Please adjust the analysis parameters and interpretations based on the specific requirements in the PDF instructions.
